---
title: "A tutorial on STOPS"
author: "Thomas Rusch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this document we introduce the functionality avalaiable in `stops` for fitting multidimensional scaling (MDS; Borg & Groenen 2005) or proximity scaling (PS) models either with a STOPS or COPS idea or not. We start with a short introduction to PS and the models that we have available. We then explain fitting of these models with the `stops` package. Next, we introduce the reader to COPS (Rusch et al. 2015a) and STOPS (Rusch et al. 2015b) models and show how to fit those. For illustration we use the `smacof::kinshipdelta` data set (Rosenberg, S. & Kim, M. P., 1975) which lists percentages of how often 15 kinship terms were not grouped together by college students. 

```{r}
library(stops)
```

## Proximity Scaling
For proximity scaling (PS) or multidimensional scaling (MDS) the input is typically an $N\times N$ matrix $\Delta^*=f(\Delta)$, a matrix of proximities with elements $\delta^*_{ij}$, that is a function of a matrix of observed non-negative dissimilarities $\Delta$ with elements $\delta_{ij}$. $\Delta^*$ usually is symmetric (but does not need to be). The main diagonal of $\Delta$ is 0. We call a $f: \delta_{ij} \mapsto \delta^*_{ij}$ a proximity transformation function. In the MDS literature these $\delta_{ij}^*$ are often called dhats or disparities. The problem that proximity scaling solves is to locate an $N \times M$ matrix $X$ (the configuration) with row vectors $x_i, i=1,\ldots,N$ in low-dimensional space $(\mathbb{R}^M, M \leq N)$ in such a way that transformations $g(d_{ij}(X))$ of the fitted distances $d_{ij}(X)=d(x_i,x_j)$---i.e., the distance between different $x_i, x_j$---approximate the $\delta^*_{ij}$ as closely as possible. We call a $g: d_{ij}(X) \mapsto d_{ij}^*(X)$ a distance transformation function. In other words, proximity scaling means finding $X$ so that $d^*_{ij}(X)=g(d_{ij}(X))\approx\delta^*_{ij}=f(\delta_{ij})$. 

This approximation $D^*(X)$ to the matrix $\Delta^*$ is found by defining a fit criterion (the loss function), $\sigma_{MDS}(X)=L(\Delta^*,D^*(X))$, that is used to measure how closely $D^*(X)$ approximates $\Delta^*$. Usually, they are closely related to the quadratic loss function. A general formulation of a loss function based on a quadratic loss is 
\begin{equation}
 \label{eq:stress}
 \sigma_{MDS}(X)=\sum^N_{i=1}\sum^N_{j=1} z_{ij} w_{ij}\left[d^*_{ij}(X)-\delta^*_{ij}\right]^2=\sum^N_{i=1}\sum^N_{j=1} z_{ij}w_{ij}\left[g\left(d_{ij}(X)\right)-f(\delta_{ij})\right]^2
\end{equation}
Here, the $w_{ij}$ and $z_{ij}$ are finite weights, with $z_{ij}=0$ if the entry is missing and $z_{ij}=1$ otherwise.  

The loss function used is then minimized to find the vectors $x_1,\dots,x_N$, i.e., 
\begin{equation}
\label{eq:optim}
\arg \min_{X}\ \sigma_{MDS}(X).
\end{equation}
There are a number of optimization techniques one can use to solve this optimization problem. 


### Stress Models 
The first popular type of PS supported in `stops` is based on the loss function type _stress_ (Kruskall 1964). This uses some type of Minkowski distance ($p > 0$) as the distance fitted to the points in the configuration, 
\begin{equation}
\label{eq:dist}
d_{ij}(X) = ||x_{i}-x_{j}||_p=\left( \sum_{m=1}^M |x_{im}-x_{jm}|^p \right)^{1/p} \ i,j = 1, \dots, N.
\end{equation}
Typically, the norm used is the Euclidean norm, so $p=2$. In standard MDS $g(\cdot)=f(\cdot)=I(\cdot)$, the identity function. 

This formulation enables one to express a large number of PS methods many of which are implemented in `stops`. In `stops` we allow to use specific choices for $f(\cdot)$ and $g(\cdot)$ from the family of power transformations so one can fit the following stress models:  

* **Explicitly normalized stress**: $w_{ij}=(\sum_{ij}\delta^{*2}_{ij})^{-1}$, $\delta_{ij}^*=\delta_{ij}$, $d_{ij}(X)^*=d_{ij}(X)$
* **Stress-1**: $w_{ij}=(\sum_{ij} d^{*2}_{ij}(X))^{-1}$, $\delta_{ij}^*=\delta_{ij}$, $d_{ij}(X)^*=d_{ij}(X)$ 
* **Sammon stress** (Sammon 1969): $w_{ij}=\delta^{*-1}_{ij}$ , $\delta_{ij}^*=\delta_{ij}$, $d_{ij}(X)^*=d_{ij}(X)$
* **Elastic scaling** stress (McGee 1966): $w_{ij}=\delta^{*-2}_{ij}$,  $\delta_{ij}^*=\delta_{ij}$, $d_{ij}(X)^*=d_{ij}(X)$
* **S-stress** (Takane et al. 1977): $\delta^*_{ij}=\delta_{ij}^2$ and $d^*_{ij}(X)=d^2_{ij}(X)$, $w_{ij}=1$
* **R-stress** (de Leeuw, 2014): $\delta^*_{ij}=\delta_{ij}$ and $d^*_{ij}=d^{2r}_{ij}$, $w_{ij}=1$  
* **Power MDS** (Buja et al. 2008, Rusch et al. 2015a): $\delta^*_{ij}=\delta_{ij}^\lambda$ and $d^*_{ij}=d^\kappa_{ij}$, $w_{ij}=1$
* **Power elastic scaling** (Rusch et al. 2015a): $w_{ij}=\delta^{*-2}_{ij}$, $\delta^*_{ij}=\delta_{ij}^\lambda$ and $d^*_{ij}=d^\kappa_{ij}$
* **Power Sammon mapping** (Rusch et al. 2015a): $w_{ij}=\delta^{*-1}_{ij}$, $\delta^*_{ij}=\delta_{ij}^\lambda$ and $d^*_{ij}=d^\kappa_{ij}$ 
* **Powerstress** (encompassing all previous models; Buja et al. 2008, Rusch et al. 2015a): $\delta^*_{ij}=\delta_{ij}^\lambda$, $d^*_{ij}=d^\kappa_{ij}$ and $w_{ij}=w_{ij}^\nu$ for arbitrary $w_{ij}$ (e.g., a function of the $\delta_{ij}$)

For all of these models one can use the function `powerStressMin` which uses majorization to find the solution (de Leeuw, 2014) . The function allows to specify a `kappa`, `lambda` and `nu` argument as well as a `weightmat` (the $w_{ij}$). 

[!!] Note that if $z_{ij}$ and $w_{ij}$ are used, then `weightmat` must be the combination of both.  

The object returned from `powerStressMin` is of class `smacofP` which extends the `smacof` classes (de Leeuw & Mair, 2009) to allow for the power transformations. Apart from that the objects are made so that they have maximum compatibility to methods from `smacof`. Accordingly, the following S3 methods are available:

| Method      | Description                   |        
|:------------|:------------------------------|   
|print        | Prints the object             |
|summary      | A summary of the object       |
|plot         | 2D Plots of the object        |
|plot3d       | Dynamic 3D configuration plot |
|plot3dstatic | Static 3D configuration plot  |
|residuals    | Residuals                     |
|coef         | Model Coefficients            | 

Let us illustrate the usage

```{r}
dis<-as.matrix(smacof::kinshipdelta)
```  

* A standard MDS (`stress`)
```{r}
res1<-powerStressMin(dis,kappa=1,lambda=1)
res1
```

* A `sammon` mapping
```{r}
res2<-powerStressMin(dis,kappa=1,lambda=1,nu=-1,weightmat=dis)
res2
```

Alternatively, one can use the faster `sammon` function from `MASS` (Venables & Ripley, 2002) for which we provide a wrapper that adds class attributes and methods (and overloads the function). 

```{r}
res2a<-sammon(dis)
res2a
```


* An `elastic` scaling
```{r}
res3<-powerStressMin(dis,kappa=1,lambda=1,nu=-2,weightmat=dis)
res3
```

* A `sstress` model 
```{r}
res4<-powerStressMin(dis,kappa=2,lambda=2)
res4
```

* An `rstress` model (with $r=1$ as $r=\kappa/2$)
```{r}
res5<-powerStressMin(dis,kappa=2,lambda=1)
res5
```

* A `powermds` model 
```{r}
res6<-powerStressMin(dis,kappa=2,lambda=1.5)
res6
```

* A `powersammon` model 
```{r}
res7<-powerStressMin(dis,kappa=2,lambda=1.5,nu=-1,weightmat=dis)
res7
```

* A `powerelastic` scaling
```{r}
res8<-powerStressMin(dis,kappa=2,lambda=1.5,nu=-2,weightmat=dis)
res8
```

* A `powerstress` model 
```{r}
res9<-powerStressMin(dis,kappa=2,lambda=1.5,nu=-1.5,weightmat=2*1-diag(nrow(dis)))
res9
summary(res9)
```

```{r,eval=FALSE,fig.show='hold',fig.width=8, fig.height = 8}
plot(res9)
plot(res9,"transplot")
plot(res9,"Shepard")
plot(res9,"resplot")
plot(res9,"bubbleplot")
```
 
[!!]: The implementation in `powerStressMin` is more a proof-of-concept than optimal. Majorizing this type of stress is a pretty hard problem and the code we use relies on a while loop in pure R. We plan to speed the loop up with a re-implementation in C in the future.

### Strain Models
The second popular type of PS supported in `stops` is based on the loss function type \emph{strain}. Here the $\Delta^*$ are a transformation of the $\Delta$, $\Delta^*= f (\Delta)$ so that $f(\cdot)=-(h\circ l)(\cdot)$ where $l$ is any function and $h(\cdot)$ is a double centering operation, $h(\Delta)=\Delta-\Delta_{i.}-\Delta_{.j}+\Delta_{..}$ where $\Delta_{i.}, \Delta_{.j}, \Delta_{..}$ are matrices consisting of the row, column and grand marginal means respectively. These then get approximated by (functions of) the inner product matrices of $X$ 
\begin{equation}
\label{eq:dist2}
d_{ij}(X) = \langle x_{i},x_{j} \rangle
\end{equation}
We can thus express classical scaling as a special case of the general PS loss with $d_{ij}(X)$ as an inner product, $g(\cdot) = I(\cdot)$ and $f(\cdot)=-(h \circ I)(\cdot)$.

If we again allow power transformations for $g(\cdot)$ and $f(\cdot)$ one can fit the following strain models with `stops` 

* **Classical scaling** (Torgerson, 1958): $\delta^*_{ij}=-h(\delta_{ij})$ and $d^*_{ij}=d_{ij}$ 
* **Powerstrain** (Buja et al. 2008, Rusch et al. 2015a): $\delta^*_{ij}=-h(\delta_{ij}^\lambda)$, $d^*_{ij}=d_{ij}$ and $w_{ij}=w_{ij}^\nu$ for arbitrary $w_{ij}$

In `stops` we have a wrapper to `cmdscale` (overloading the `base` function) which extend functionality by offering an object that matches `smacofP` objects with corresponding methods.

A `powerstrain` model is rather easy to fit with simply subjecting the dissimilarity matrix to some power. Here we use $\lambda=3$.

```{r}
resc<-cmdscale(kinshipdelta^3)
resc
```

```{r}
summary(resc)
```


```{r,eval=FALSE,fig.show='hold',fig.width=8, fig.height=8}
summary(resc)
plot(resc)
```

## Augmenting MDS with structure considerations: STOPS and COPS 

The main contribution of the `stops` package is not in solely fitting the _powerstress_ or _powerstrain_ models and their variants from above, but allowing to choose the right transformation to achieve a "structured" MDS result automatically. This can be useful in a variety of contextes: to explore or generate structures, to restrict the target space, to avoid artefacts, to preserve certain types of structures and so forth. 

For this, an MDS loss function is subjected to nonlinear transformations and is augmented to include penalties for the type of structures one is aiming for. This combination of an MDS loss with a structuredness penalty is what we call "structure optimized loss" (stoploss) and the resulting MDS is coined "Structured Optimized Proximity Scaling" (or STOPS). The prime example for a STOPS model is "Cluster Optimized Proximity Scaling" (COPS) which selects optimal parameters so that the clusteredness appearance of the configuation is improved (see below). 

### STOPS

Following Rusch et al. (2015b) the general idea  is that from given observations $\Delta$ we look for a configuration $X$. This achieves this by minimizing some loss function $\sigma_{MDS}(X^*;\Delta^*)$ where the $\Delta^*, X^*$ are functions of the $\Delta$ and $X$. The $X$ has properties with regards to its structural appearance, which we call _c-structuredness_ for configuration-structuredness. There are different types of c-structuredness people might be interested in (say, how clustered the result is, or that dimensions are orthogonal or if there is some submanifold that the data live on). We developed indices for these types of c-structuredness that capture that essence in the configuration.

We have as part of a STOPS model a proximity scaling loss function $\sigma_{MDS}(\cdot)$, a $\Delta$ and an $X$ and some transformation $f_{ij}(\delta_{ij};\theta)$ and $g_{ij}(d_{ij};\theta)$ that is parametrized (with $\theta$ either finite or infinite dimensional, e.g., a transformation applied to all observations like a power transformation or even an individual transformation per object). These transformations achieve a sort of push towards more structure, so different values for $\theta$ will in general lead to different c-structuredness.

We further have $K$ different indices $I_k(X^*)$ that measure different types of c-structuredness. We can then define \emph{STOPS} as methods that are of the form (additive STOPS)
\begin{equation}
  \text{aSTOPS}(X, \theta, v_0, \dots, v_k; \Delta) = v_0 \cdot \sigma_{MDS}(X^*(\theta)) + \sum^K_{k=1} v_k I_k(X^*(\theta))
\end{equation}
 or (multiplicative STOPS)
\begin{equation}
  \text{mSTOPS}(X, \theta, v_0, \dots, v_k; \Delta) = \sigma_{MDS}(X^*(\theta))^{v_0} \cdot \prod^K_{k=1} I_k(X^*(\theta))^{v_k} 
\end{equation}
(which can be expressed as aSTOPS by logarithms). Here the $v_0,...,v_k$ are weights that determine how the individual parts (mds loss and c-structuredness indices) are aggregated. 

[//]: Note that in this formulation the aggregation is a sum/product, so the weights must be negative if the mds loss should be penalized and a higher index stands for more structure. Alternatively, the `stressweight` can be negative and the `strucweight` positive. 

The job is then to find 
\begin{equation}  
 \arg\min_{\vartheta}\ \text{aSTOPS}(X, \theta, v_0, \dots, v_k; \Delta)\ \text{or} \ \arg\min_{\vartheta}\ \text{mSTOPS}(X, \theta, v_0, \dots, v_k; \Delta)
\end{equation} 
where $\vartheta \subseteq \{X,\theta, v_0, \dots, v_k\}$.  Typically $\vartheta$ will be a subset of all possible parameters here (e.g., the weights might be given _a priori_). Currently, the transformations that can be used in `stops` are limited to power transformations. 

Minimizing stoploss can be difficult. In `stops` we use a nested algorithm combining optimization that internally first solves for $X$ given $\theta$, $\arg\min_X \sigma_{MDS}\left(X,\theta\right)$, and then optimize over $\theta$ with a metaheuristic. Implemented are a simulated annealing (`optimmethod="SANN"`) or particle swarm optimization (`optimmethod="pso"`) and a variant of the Luus-Jaakola (`optimmethod="ALJ"`) procedure \citep{luus1973optimization}. We suggest to use the latter.

Currently the following c-structuredness types are supported:

* **c-clusteredness** (`cclusteredness`): A clustered appearance of the configuration ($I_k$ is the normed OPTICS cordillera (COPS; Rusch et al. 2015a); 0 means no c-clusteredness, 1 means perfect c-clusteredness)
* **c-linearity** (`clinearity`): Projections lie close to a linear subspace of the configuration ($I_k$ is maximal multiple correlation; 0 means orthogonal, 1 means perfectly linear) 
* **c-manifoldness** (`cmanifoldness`): Projections lie on a sub manifold of the configuration ($I_k$ is maximal correlation (Sarmanov, 1958); only available for two dimensions; 1 means perfectly smooth function)
* **c-dependence** (`cdependence`): Random vectors of projections onto the axes are stochastically dependent ($I_k$ is distance correlation (Szekely et al., 2007); only available for two dimensions; 0 means they are stochastically independent)
* **c-association** (`cassociation`): Pairwise nonlinear association between dimensions ($I_k$ is the pairwise maximal maximum information coefficient (Reshef et al. 2011), 1 means perfect functional association)  
* **c-nonmonotonicity** (`cnonmonotonicity`): Deviation from monotonicity ($I_k$ is the pairwise maximal maximum assymmetry score (Reshef et al. 2011), the higher the less monotone)  
* **c-functionality** (`cfunctionality`): Pairwise functional, smooth, noise-free relationship between dimensions ($I_k$ is the mean pairwise maximum edge value (Reshef et al. 2011), 1 means perfect functional association)
* **c-complexity** (`ccomplexity`): Measures the degree of complexity of the functional relationship between any two dimensions ($I_k$ is the pairwise maximal minimum cell number (Reshef et al. 2011), the higher the more complex) 
* **c-faithfulness** (`cfaithfulness`): How accurate is the neighbourhood of $\Delta$ preserved in $D$ ($I_k$ is the adjusted Md index of Chen \& Buja, 2013; note that this index deviates from the others by being a function of both $X^*$ and $\Delta^*$ rather than $X^*$ alone)
* c-randomness: How close to a random pattern (under some model) is the configuration ($I_k$ is not clear yet; not yet implemented)
[//]: * c-evenness: How even and/or spread out are observations ($I_k$ is a function of the Ripley's K and L functions; not yet implemented) * c-randomness: How close to a random pattern (under some model) is the configuration ($I_k$ is not clear yet; not yet implemented) * c-separability: How well are classes (linear and nonlinear) separable. Perhaps only useful if there are labels. ($I_k$ is a separation index; not yet implemented) * c-sparsity: How sparse is the representation (percentage of vectors with 0 or percentage of weights that are zero; not yet implemented) * c-optimality: Finding the global minimum of stress by adding a penalty when the optimum is local ($I_k$ is Jan's idea; not yet implemented)

If we have a single $I(X)=OC(X)$, the OPTICS cordillera (Rusch et al. 2015a), and the transformations applied are power transformations and the weights for the $I(X)$ is negative we essentially have COPS (see below).    

For the MDS loss (argument `loss` in functions `stops` and `cops`), the functions currently support all losses derived from _powerstress_ and _powerstrain_ and can in principle be fitted with `powerStressMin` alone. However, for many models offer dedicated functions that either use workhorses that are more optimized for the problem at hand and/or restrict the parameter space for the distance/proximity transformations and thus can be faster. They are: 

* `stress`, `smacofSym`: Kruskall's stress; Workhorse: `smacofSym`, Optimization over $\lambda$
* `smacofSphere`: Kruskall's stress for projection onto a sphere; Workhorse `smacofSphere`, Optimizes over $\lambda$
* `strain`, `powerstrain`: Classical scaling; Workhorse: `cmdscale`, Optimization over $\lambda$
* `sammon`, `sammon2`: Sammon scaling; Workhorse: `sammon` or `smacofSym`, Optimization over $\lambda$
* `elastic`: Elastic scaling; Workhorse: `smacofSym`, Optimization over $\lambda$
* `sstress`: S-stress; Workhorse: `powerStressMin`, Optimization over $\lambda$ 
* `rstress`: S-stress; Workhorse: `powerStressMin`, Optimization over $\kappa$
* `powermds`: MDS with powers; Workhorse: `powerStressMin`, Optimization over $\kappa$, $\lambda$
* `powersammon`: Sammon scaling with powers; Workhorse: `powerStressMin`, Optimization over $\kappa$, $\lambda$
* `powerelastic`: Elastic scaling with powers; Workhorse: `powerStressMin`, Optimization over $\kappa$, $\lambda$
* `powerstress`: Power stress model; Workhorse: `powerStressMin`, Optimization over $\kappa$, $\lambda$, $\nu$
 
#### Usage 

The syntax for fitting a `stops` model is rather straightforward. One has to supply the arguments `dis` which is a dissimilarity matrix and `structures` a character vector listing the c-structuredness type that should be used to augment the PS loss (see above for the types of structures and losses). The parameters for the structuredness indices should be given with `strucpars`, a list whose elements are lists corresponding to each structuredness index and listing the parameters. The PS loss can be chosen with the argument `loss`. The type of aggregation for the multi-objective optimization is specified in `type` and can be one of `additive` or `multiplicative`. One can pass additional parameters to the fitting workhorses with `...`.

`stops(dis, structures = c("cclusteredness","clinearity"), loss="stress", ...)`

One then has all the S3 methods of `smacofP` at one's disposal.

For example, let us fit an mSTOPS model that looks for a transformation of the $\delta_{ij}$ so that a) the result has maximal c-clusteredness (which is 1 in the best case, so we set a negative weight for this structure) b) the projection onto the principal axes are nearly orthogonal (c-clinearity close to 0, so we set a positive weight for this structure) c) but the projections onto the principal axes should be stochastically dependent (negative weight on c-dependence) and d) the fit of the MDS is also factored in (so we set positive weight on the MDS loss). Since we use mSTOPS and a negative weight for c-linearity and c-dependence, a c-linearity/c-dependence close to 0 will overall dominate the stoploss function with the other two criteria being more of an afterthought - in aSTOPS that would be different. 

[!!]: This is generally the approach to be chosen: We _minimize_ the stoploss, so a c-structuredness index that should be (numerically) large needs a _negative weight_ and a c-structuredness index that should be (numerically) small needs a _positive weight_.  

We first set up the parameters for the structuredness indices. For the OPTICS cordillera we use a $d_max$ of 1, eps=10 and minpts=2, for c-linearitry we have noo parameters (so using `NULL` will work) and for the c-dependence we have a single parameter, `index`, which we set to 2.   

```{r}
strucpars<-list(list(eps=10,minpts=2,rang=c(0,1)), #cordillera
                NULL,     # c-linearity (has no parameters)
                list(c(index=2) #c-dependence
                ) 

```

```{r}
ressm<-stops(kinshipdelta,loss="stress",stressweight=1,structures=c("cclusteredness","clinearity","cdependence"),strucweight=c(-0.33,0.33,-0.33),verbose=0,strucpars=strucpars,type="multiplicative")
ressm
```

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(ressm)
```

Let us compare this with the corresponding aSTOPS

```{r}
ressa<-stops(kinshipdelta,loss="stress",stressweight=1,structures=c("cclusteredness","clinearity","cdependence"),strucweight=c(-0.33,0.33,-0.33),verbose=0,strucpars=strucpars,type="additive")
ressa
```

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(ressa)
```

We see that the c-clusteredness is higher as compared to the mSTOPS result - we have a number of distinct object clusters (with at least minpts=2) and they are more spread out and distributed more evenly. The dimensions on the other hand are now farther from being orthogonal but the stochastic dependence is higher (which is a non-linear one obviously). 

For convenience it is also possible to use the `stops` function for finding the loss-optimal transformation in the the non-augmented models specified in `loss`, by setting the `strucweight`, the weight of any c-structuredness, to 0. Then the function optimizes the MDS loss function only. 

```{r,eval=FALSE}
ressa<-stops(kinshipdelta,structure=c("clinearity"),strucweight=0,loss="stress",verbose=0)
```

### COPS

A special STOPS model is COPS (Rusch et al. 2015a) for "Cluster Optimized Proximity Scaling". This is also one of the main use cases for STOPS models. Let us write $X(\theta)=\arg\min_X \sigma_{MDS}(X,\theta)$ for the optimal configuration for given transformation parameter $\theta$. Following the outline of STOPS the overall objective function, which we call \emph{cluster optimized loss (coploss)}, is a weighted combination of the $\theta-$parametrized loss function, $\sigma_{MDS}\left(X(\theta),\theta\right)$, and a c-clusteredness measure, the OPTICS cordillera or $OC(X(\theta);\epsilon,k,q)$ to be optimized as a function of $\theta$ or
\begin{equation}
\label{eq:spstress}
\text{coploss}(\theta) = v_1 \cdot \sigma_{MDS}\left(X(\theta),\theta \right) - v_2 \cdot \text{OC}\left(X(\theta);\epsilon,k,q\right) 
\end{equation}
with $v_1,v_2 \in \mathbb{R}$ controlling how much weight should be given to the scaling fit measure and the c-clusteredness. In general $v_2,v_2$ are either \emph{a priori} determined values that make sense for the application or may be used to trade-off fit and c-clusteredness in a way for them to be commensurable. In the latter case we suggest taking the fit function value as it is ($v_1=1$) and fixing the scale such that $\text{coploss}=0$ for the scaling result with no transformations ($\theta=\theta_0$), i.e.,
\begin{equation}
\label{eq:spconstant0}
v^{0}_{1}=1, \quad v^{0}_2=\frac{\sigma_{MDS}\left(X(\theta_0),\theta_0\right)}{\text{OC}\left(X(\theta_0);\epsilon,k,q\right)},
\end{equation}
with $\theta_0=(1,1)^\top$ in case of loss functions with power transformations. Thus an increase of 1 in the MDS loss measure can be compensated by an increase of $v^0_1/v^0_2$ in c-clusteredness. Selecting $v_1=1,v_2=v^{0}_2$ this way is in line with the idea of pushing the configurations towards a more clustered appearance relative to the initial solution. 

 Another possibility is to choose them in such a way that $\text{coploss}=0$ in the optimum value, i.e., choosing $v^{opt}_{1}, v^{opt}_2$ so that 
  \begin{equation}
v^{opt}_1 \cdot \sigma_{MDS}\left(X(\theta^*),\theta^*\right)-v^{opt}_2 \cdot \text{OC}\left(X(\theta^*);\epsilon,k,q \right) = 0 
\end{equation}
with $\theta^*:=\arg\min_\theta \text{coploss}(\theta)$. This is in line with having $\text{coploss}(\theta)>0$ for $\theta \neq \theta^*$ and allows to optimize over $v_1,v_2$. 

The optimization problem in COPS is then to find    
\begin{equation}
\label{eq:soemdsopt2}
\arg\min_{\theta} \text{coploss}(\theta)
\end{equation}
by doing
\begin{equation} 
\label{eq:soemdsopt}
v_1 \cdot \sigma_{MDS}\left(X(\theta),\theta\right) - v_2 \cdot \text{OC}\left(X(\theta);\epsilon,k,q\right) \rightarrow \min_\theta! 
\end{equation}
For a given $\theta$ if $v_2=0$ than the result of optimizing the above is the same as solving the respective original PS problem. Letting $\theta$ be variable, $v_2=0$ will minimize the loss over configurations obtained from using different $\theta$.

[//]: Minimizing coploss is difficult. In `stops` we use a nested algorithm combining optimization that internally first solves for $X$ given $\theta$, $\arg\min_X \sigma_{MDS}\left(X,\theta\right)$, and then optimize over $\theta$ with a metaheuristic. Implemented are a simulated annealing (`optimmethod="SANN"`) or particle swarm optimization (`optimmethod="pso"`) and a variant of the Luus-Jaakola (`optimmethod="ALJ"`) procedure \citep{luus1973optimization} to be used in Step 3 that usually converges in less than 200 iterations to an acceptable solution. We suggest to use the latter.

The c-clusteredness index we use is the OPTICS cordillera which measures how clustered a configuration appears. It is based on the OPTICS algorithm that outputs an ordering together with a distance. The OPTICS cordillera is now simply an agregation of that information. Since we know what constitutes a maximally clustered result, we can derive an upper bound and normalize the index to lie between 0 and 1. If it is maximally clustered, the index gets a value of 1,and it gets 0 if all points are equidistant to their nearest neighbours (a matchstick embedding). See Rusch et al (2015a) for details. 

#### Usage 
Even though one can fit a COPS model with `stops`, there is a dedicated function `cops`. Its syntax works pretty much like in `stops` only that the `structure` argument is non-existant. 

`cops(dis,loss,...)`

For the example we use a COPS model for a classical scaling (`strain` loss) 

```{r}
resc<-cops(kinshipdelta,loss="strain")
resc
summary(resc)
```
A number of plots are availabe

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(resc,"confplot")
plot(resc,"Shepard")
plot(resc,"transplot")
plot(resc,"reachplot")
````

For convenience it is also possible to use the `cops` function for finding the loss-optimal transformation in the the non-augmented models specified in `loss`, by setting the `cordweight`, the weight of the OPTICS cordillera, to 0. Then the function optimizes the MDS loss function only. 

```{r}
resca<-cops(kinshipdelta,cordweight=0,loss="strain")
resca
```

Here the results match the result from using the standard `cordweight`. We can give more weight to the c-clusteredness though:

```{r}
rescb<-cops(kinshipdelta,cordweight=20,loss="strain")
rescb
```

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(resca,main="with cordweight=0")
plot(rescb,main="with cordweight=20")
```

This result has more c-clusteredness but less fit. The higher c-clusteredness is discernable in the Grandfather/Brother and Grandmother/Sister clusters (we used a minimum number of 2 observations to make up a cluster, `minpts=2`). 

## Other Functions
The package also provides functions that are used by the `cops` and `stops` and `powerStressMin` functions but may be of interest to a end user beyond that.

### OPTICS and OPTICS cordillera
For calculating a COPS solution, we need the OPTICS algorithm and the OPTICS cordillera. In the package we also provide a rudimentary interface to the OPTICS impementation in ELKI. 

```{r}     
data(iris)
res<-optics(iris[,1:4],minpts=2,epsilon=1000)
print(res)
summary(res)
```

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(res,withlabels=TRUE)
```

and a function for calculating and displaying the OPTICS cordillera.      

```{r,fig.show='hold',fig.width=8,fig.height=8}
cres<-cordillera(iris[,1:4],minpts=2,epsilon=1000,scale=FALSE)
cres
summary(cres)
```

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(cres)
```

### Optimization
Since the inner optimization problem in STOPS models is hard and takes long, Rusch et al. (2015a) developed a metaheuristic for the outer optimization problem that needs typically less calls to the inner minimization than `pso` or `SANN`, albeit without the guarantees of convergence to a global minimum for non-smooth functions. It is an adaptation of the Luus-Jaakola random search (Luus & Jaakola 1973). It can used with the function `ljoptim` which modeled its output after `optim`. It needs as arguments `x` a starting value, `fun` a function to optimize, a `lower` and `upper` box constraint for the search region. By using the argument `adaptive=TRUE` or `FALSE` one can switch between our adaptive version and the original LJ algorithm. Accuracy of the optimization can be controlled with the `maxit` (maximum number of iterations), `accd` (terminates after the length of the search space is below this number ) and `acc` arguments (terminates if difference of two subsequent function values are below this value).  

We optimize a "Wild Function" with the non-adaptive LJ version (and numerical accuracies of at least `1e-16` for `accd` and `acc`).
 
```{r}
set.seed(210485)
fwild <- function (x) 10*sin(0.3*x)*sin(1.3*x^2) + 0.00001*x^4 + 0.2*x+80
res2<-ljoptim(50, fwild,lower=-50,upper=50,adaptive=FALSE,accd=1e-16,acc=1e-16)
res2
```

```{r,fig.show='hold',fig.width=8,fig.height=8}
plot(fwild, -50, 50, n = 1000, main = "ljoptim() minimising 'wild function'")
points(res2$par,res2$value,col="red",pch=19)
```    

### Procrustes Adjustment
We also crovide a procrustes adjustemnt to make to configuration visually comparable. The function is `conf_adjust` and takes two configurations `conf1` the reference configuration and `conf2` another configuration. It returns the adjusted versions

```{r,eval=FALSE}
conf_adjust(conf1,conf2) 
```


##References

* Borg I, Groenen PJ (2005). Modern multidimensional scaling:  Theory and applications.  2nd edition. Springer, New York

* Buja A, Swayne DF, Littman ML, Dean N, Hofmann H, Chen L (2008). Data visualization with multidimensional scaling. Journal of Computational and Graphical Statistics, 17 (2), 444-472.

* Chen L, Buja A (2013). Stress functions for nonlinear dimension reduction, proximity analysis, and graph drawing. Journal of Machine Learning Research, 14, 1145-1173.

* de Leeuw J (2014). Minimizing r-stress using nested majorization. Technical Report, UCLA, Statistics Preprint Series.

* de Leeuw J, Mair P (2009). Multidimensional Scaling Using Majorization:  SMACOF in R. Journal of Statistical Software, 31 (3), 1-30. 

* Kruskal JB (1964).  Multidimensional scaling by optimizing goodness of fit to a nonmetric hypothesis. Psychometrika, 29 (1), 1-27.

* Luus R, Jaakola T (1973).  \Optimization by direct search and systematic reduction of the size of search region. American Institute of Chemical Engineers Journal (AIChE), 19 (4), 760-766.

* McGee VE (1966). The multidimensional analysis of 'elastic' distances. British Journal of Mathematical and Statistical Psychology, 19 (2), 181-196.

* Reshef D, Reshef Y, Finucane H, Grossman S, McVean G, Turnbaugh P, Lander E, Mitzenmacher M, Sabeti P (2011). Detecting novel associations in large datasets. Science, 334, 6062.

* Rosenberg, S. & Kim, M. P. (1975). The method of sorting as a data gathering procedure in multivariate research. Multivariate Behavioral Research, 10, 489-502.

* Rusch, T., Mair, P. and Hornik, K. (2015a) COPS: Cluster Optimized Proximity Scaling. Discussion Paper Series / Center for Empirical Research Methods, 2015/1. WU Vienna University of Economics and Business, Vienna.

* Rusch,  T.,  Mair,  P. and Hornik, K. (2015b). Structuredness  Indices  and  Augmented  Nonlinear Dimension Reduction. In preparation.

* Sammon JW (1969). A nonlinear mapping for data structure analysis. IEEE Transactions on Computers, 18 (5), 401-409

* Sarmanov OV (1958) The maximum correlation coefficient (symmetric case). Dokl. Akad. Nauk SSSR, 120 : 4 (1958), 715 - 718.

* Székely, G. J. Rizzo, M. L. and Bakirov, N. K. (2007). Measuring and testing independence by correlation of distances, The Annals of Statistics, 35:6, 2769–2794.

* Takane Y, Young F, de Leeuw J (1977). Nonmetric individual differences multidimensional scaling: an alternating least squares method with optimal scaling features. Psychometrika, 42 (1), 7-67.

* Torgerson WS (1958). Theory and methods of scaling. Wiley.


* Venables WN, Ripley BD (2002). Modern Applied Statistics with S. Fourth edition. Springer, New York. 




